---
title: "Skull Scoring"
author: "Nicole Ackermans"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#last updated 12 Nov 2024

#library
```{r}
library(readxl) # read excel files
library(tidyverse)
library(ggpubr) # annotate graphs
library(effectsize) # Chisq effect size calculation
library(janitor) # clean names
library(santoku) # chop quantiles
```

#import files
```{r}
df <- read_excel("Skull Scoring_Niki_240423_EB.xlsx")
df
```

#clean up
```{r}
df <- df |>
  drop_na(species) |>
  separate(species, c("species", "id"), ":") |>
  mutate(rb_joy = (`RED- Joy` + `BLUE-Joy`),
         rb_niki = (`RED- Niki` + `BLUE- Niki`),
         rb_emma = (`RED- Emma` + `BLUE- Emma`),
         rb_lillie = (`RED- Lillie` + `BLUE- Lillie`),
         rb_maren = (`RED- Maren` + `BLUE- Maren`)) |>
  rename(voxel.size = `Voxel Size`) |>
  rowwise() |>
  mutate(ERI = mean(c(rb_joy, rb_niki, rb_emma, rb_lillie, rb_maren)))

df
  
```

#roughness categories & average when multiple individuals
```{r}
df <- df |> 
  group_by(species) |>
  summarise(headbutting = first(headbutting),
            weight = first(weight),
            meanERI = mean(ERI),
            meanVoxel = mean(voxel.size)) |>
  mutate(ERI_cat = case_when(meanERI >= median(meanERI) ~ "high",
                               .default = "low")) |>
  mutate(headbutting = as.factor(headbutting),
         ERI_cat = as.factor(ERI_cat))
df

```

#test for normal distribution
```{r, fig.width = 8, fig.height = 4}

p1 <- ggdensity(df$meanVoxel) +
   labs(x = "mean voxel size")

p2 <- ggdensity(df$weight) +
  labs(x = "weight")

p3 <- ggdensity(df$meanERI) +
  labs(x = "mean ERI score")

ggarrange(p1, p2, p3, nrow = 1)

#if p > 0.05 the distribution of the data are not significantly different from 
#normal distribution. We can assume normality
shapiro.test(df$meanVoxel)
shapiro.test(df$weight)
shapiro.test(df$meanERI)
```

#test for normal distribution in the headbutting group
```{r, fig.width = 8, fig.height = 4}
df_hb <- df |>
  filter(headbutting == "Y")

p1 <- ggdensity(df_hb$meanVoxel) +
   labs(x = "mean voxel size")

p2 <- ggdensity(df_hb$weight) +
  labs(x = "weight")

p3 <- ggdensity(df_hb$meanERI) +
  labs(x = "mean ERI score")

ggarrange(p1, p2, p3, nrow = 1)

#if p > 0.05 the distribution of the data are not significantly different from 
#normal distribution. We can assume normality
shapiro.test(df_hb$meanVoxel)
shapiro.test(df_hb$weight)
shapiro.test(df_hb$meanERI)
```

#test for normal distribution in the non-headbutting group
```{r, fig.width = 9, fig.height = 3}
df_nohb <- df |>
  filter(headbutting == "N")

p1 <- ggdensity(df_nohb$meanVoxel) +
   labs(x = "mean voxel size")

p2 <- ggdensity(df_nohb$weight) +
  labs(x = "weight")

p3 <- ggdensity(df_nohb$meanERI) +
  labs(x = "mean ERI score")

distribution <- ggarrange(p1, p2, p3, nrow = 1)
distribution
#if p > 0.05 the distribution of the data are not significantly different from 
#normal distribution. We can assume normality
shapiro.test(df_nohb$meanVoxel)
shapiro.test(df_nohb$weight)
shapiro.test(df_nohb$meanERI)

#ggsave("distribution_241108.png", distribution, dpi = 300)
```

#bivariate plot CT quality vs endocranial roughness faceted by headbutting
```{r, fig.width = 6, fig.height = 3}
bivariate <- df |>
  ggplot(aes(x = meanVoxel, y = meanERI)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  stat_cor(method = "spearman", label.x = 0.7, label.y = 15, cor.coef.name = "Rho") +
  geom_text(aes(label = species), size = 2) +
  facet_grid(cols = vars(headbutting)) +
  labs(title = "faceted by headbutting Y/N")
  
bivariate

#ggsave("bivariate_241108.png", bivariate, dpi = 300)
  
```

#bivariate plot weight vs endocranial roughness
```{r}
df |>
  ggplot(aes(x = weight, y = meanERI)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  stat_cor(method = "spearman", label.x = 400, label.y = 15, cor.coef.name = "Rho") +
  geom_text(aes(label = species), size = 3) 
```

#Chi square test headbutting vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
comparing two binary factors (yes/no)

Headbutting is categorical, therefore ERI must be categorical in order to compare them
```{r}

chisq <- chisq.test(df$headbutting, df$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```

#Chi-squared test CT quality (assessed by mean voxel size) vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
#make categorical
df <- df |> 
  mutate(voxel_cat = chop_quantiles(meanVoxel, c(0, 0.25, 0.50, 0.75),
                                   labels = c("small", "medium", "big", "bigger"))
         )


#chi-squared test
chisq <- chisq.test(df$voxel_cat, df$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```

#Chi square test mean Voxel vs weight
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
#make categorical
df <- df |> 
  mutate(weight_cat = chop_quantiles(meanVoxel, c(0, 0.25, 0.50, 0.75),
                                   labels = c("small", "medium", "big", "bigger"))
         )

chisq <- chisq.test(df$voxel_cat, df$weight_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```
# compare within lower weight animals
```{r}
df_light <- df |>
  filter(weight < 60)
df_light
```
#Chi square test headbutting vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
comparing two binary factors (yes/no)
```{r}
chisq <- chisq.test(df_light$headbutting, df_light$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```

#Chi square test CT quality vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df_light$voxel_cat, df_light$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v

```
# compare within heavier weight animals
```{r}
df_heavy <- df |>
  filter(weight > 60)
df_heavy
```
#Chi square test headbutting vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
comparing two binary factors (yes/no)
```{r}
chisq <- chisq.test(df_heavy$headbutting, df_heavy$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```

#Chi square test CT quality vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df_heavy$voxel_cat, df_heavy$ERI_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v

```
#Chi square test CT quality vs weight
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df_heavy$voxel_cat, df_heavy$weight_cat)

chisq

n <- sum(chisq$observed)  # Total number of observations
nrow <- nrow(chisq$observed)  # Number of rows in contingency table
ncol <- ncol(chisq$observed)  # Number of columns in contingency table

# Calculate Cramer's V with confidence interval
cramers_v <- chisq_to_cramers_v(
  chisq = chisq$statistic,
  n = n,
  nrow = nrow,
  ncol = ncol,
  adjust = TRUE,
  ci = 0.95,
  alternative = "greater"
)

cramers_v
```

#bivariate plot headbutting vs endocranial roughness at heavy vs light
```{r, fig.width = 10, fig.height = 5}

light <- df_light |>
  ggplot(aes(x = meanVoxel, y = meanERI)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  stat_cor(method = "spearman", label.x = 0.5, label.y = 15, cor.coef.name = "Rho") +
  geom_text(aes(label = species), size = 3) +
  labs(title = "species < 60 kg")

heavy <- df_heavy |>
  ggplot(aes(x = meanVoxel, y = meanERI)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  stat_cor(method = "spearman", label.x = 1, label.y = 15, cor.coef.name = "Rho") +
  geom_text(aes(label = species), size = 3) +
  labs(title = "species > 60 kg")

ggarrange(light, heavy)
  
```

# boxplot
```{r, fig.width = 3, fig.height = 3}

df |>
  mutate(headbutting = ifelse(headbutting == "N", "non-headbutting", 
                              ifelse(headbutting == "Y", "headbutting", 
                                     headbutting))) |>
  ggplot(aes(x = headbutting, y = meanERI, fill = headbutting)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.4) +
  theme_pubclean() +
  theme(axis.title.x = element_blank(),   
        legend.position = "none") 

#ggsave("HbBoxplot_241107.png", plot1, dpi = 300)
```

