---
title: "Skull Scoring"
author: "Nicole Ackermans"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#library
```{r}
library(readxl) # read excel files
library(tidyverse)
library(ggpubr) # annotate graphs
library(effectsize) #Chisq effect size calculation
```

#import files
```{r}
df <- read_excel("Skull Scoring_Niki_240423.xlsx")
df
```

#clean up
```{r}
df <- df |>
  drop_na(species) |>
  separate(species, c("species", "id"), ":") |>
  mutate(rb_joy = (`RED- Joy` + `BLUE-Joy`),
         rb_niki = (`RED- Niki` + `BLUE- Niki`),
         rb_emma = (`RED- Emma` + `BLUE- Emma`),
         rb_lillie = (`RED- Lillie` + `BLUE- Lillie`),
         rb_maren = (`RED- Maren` + `BLUE- Maren`)) |>
  rowwise() |>
  mutate(ERI = mean(c(rb_joy, rb_niki, rb_emma, rb_lillie, rb_maren)))

df
  
```
#roughness categories & average when multiple individuals
```{r}
df <- df |> 
  group_by(species) |>
  summarise(headbutting = first(headbutting),
            CTquality = first(CTquality),
            weight = first(weight),
            meanERI = mean(ERI)) |>
  mutate(ERI_cat = case_when(meanERI >= median(meanERI) ~ "high",
                               .default = "low")) |>
  mutate(headbutting = as.factor(headbutting),
         ERI_cat = as.factor(ERI_cat))
df

```
#Chi square test headbutting vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
comparing two binary factors (yes/no)
```{r}

chisq <- chisq.test(df$headbutting, df$ERI_cat)
chisq

# Calculate Cramer's V with confidence interval
effectsize(chisq, adjust = TRUE)
```
#t test headbutting vs endocranial roughness score
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
a binary factor (yes/no) to a continuous variable (score)
```{r}
t.test(meanERI ~ headbutting, data = df)
```

#Chi square test CT quality vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df$CTquality, df$ERI_cat)
chisq

# Calculate Cramer's V with confidence interval
effectsize(chisq, adjust = TRUE)
```
#Chi square test CT quality vs weight
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df$CTquality, df$weight)
chisq

# Calculate Cramer's V with confidence interval
effectsize(chisq, adjust = TRUE)
```
# compare within similar weight animals
```{r}
df_sameweight <- df |>
  filter(weight > 60)
df_sameweight
```
#Chi square test headbutting vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
comparing two binary factors (yes/no)
```{r}
chisq <- chisq.test(df_sameweight$headbutting, df_sameweight$ERI_cat)
chisq

# Calculate Cramer's V with confidence interval
effectsize(chisq, adjust = TRUE)

```

#t test headbutting vs endocranial roughness score
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
a binary factor (yes/no) to a continuous variable (score)
```{r}
t.test(meanERI ~ headbutting, data = df_sameweight)
```

#Chi square test CT quality vs endocranial roughness
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df_sameweight$CTquality, df_sameweight$ERI_cat)
chisq

# Calculate Cramér's V with confidence interval
effectsize(chisq, adjust = TRUE)
```

#Chi square test CT quality vs weight
If p > 0.05 we would say the two categories are independent (i.e., there is no association)
```{r}
chisq <- chisq.test(df_sameweight$CTquality, df_sameweight$weight)
chisq

# Calculate Cramér's V with confidence interval
effectsize(chisq, adjust = TRUE)
```

# boxplots
```{r, fig.width = 3, fig.height = 3}
plot1 <- df |>
  mutate(headbutting = ifelse(headbutting == "N", "non-headbutting", 
                              ifelse(headbutting == "Y", "headbutting", 
                                     headbutting))) |>
  ggplot(aes(x = headbutting, y = meanERI, fill = headbutting)) +
  geom_boxplot() +
  theme_pubclean() +
  theme(axis.title.x = element_blank(),   
        legend.position = "none") 

plot1

#ggsave("HbBoxplot_240723.png", plot1, dpi = 300)
```

